A Decision Procedure for (Co)datatypes in SMT Solvers

### 2014-12-22 tex/codatadp.tex
minor

\hyphenation{data-type data-types co-data-type co-data-types isa-belle sledge-hammer}

@@

\paragraph{Correctness.}
An application of a rule is \emph{redundant} if at least one branch in the
conclusion does not add a new equality to $\Ec$ and is not $\bot$.
%%% @ANDY: Is the definition or redundant really OK? Or should it read "if none of the branches add a new eq. or is \bot"?
A \emph{derivation tree} is a tree whose nodes are finite sets of equalities, where non-root nodes are obtained by
a nonredundant application of a derivation rule to its parent node.
A derivation tree is \emph{closed} if all of its leaf nodes are $\bot$.
A node is \emph{saturated} if no nonredundant instance of a rule can be applied to it.
%
If there exists a closed derivation tree with root node $\Ec$, we conclude that $\Ec$ is $\thD$-unsatisfiable.
If there exists a derivation tree with root node $\Ec$ that contains a saturated node, we conclude that $\Ec$ is $\thD$-satisfiable.

@@

We now define the class of interpretations for $\thD$.
In the following, $\M( \tau )$ denotes the interpretation type $\tau$ in $\M$,
that is, a set of domain elements for that type.
For a term $t$, $\M( t )$ denotes the interpretation of $t$ in $\M$,
that is, an element of $\M( \tau )$ where $\tau$ is the type of $t$.
In the following,
\emph{all} types are interpretated as a set of $\mu$-terms.
This is done to simplify the presentation.
Since all values of types in $\Data$ and $\Nondata$ are acyclic,
a typical representation for these values can be obtained by
dropping each of its $\mu$-bindings. 
For instance, the $\mu$-term $\mu x. \const{0}$ representing the integer constant zero
is simply $\const{0}$ after dropping the $\mu$-binding of $x$.

@@

We now show that if statements 2 and 3 hold for some $\Val_1$,
then 2 and 3 also hold for $\nf{(\Val_1 \sigma)}$,
where $\sigma$ is a substitution of the form $\{ {\Varec{x}} \mapsto \mu {\Varec{x}}. t \}$,
${\Varec{x}} \in FV( \Val_1 )$,
and $\mu {\Varec{x}}. t$ is not $\alpha$-equivalent to a term in $\mathcal{T}^\M_{\Varec{x}}(\Val_1)$.
To show 2,
by assumption of 2 on $\Val_1$, we have $\Val_1 \ec{y} \interp{ u } = \Val_1 \ec{x}$ for
all subterms $u$ of $\Val_1 \ec{y}$ where $\muvar(t) = {\Varec{x}}$.
Thus, $\Val_1 \sigma \ec{y} \interp{ u } = \Val_1 \sigma \ec{x}$ and
$\nf{(\Val_1 \sigma)} \ec{y} \interp{ u } = \nf{(\Val_1 \sigma)} \ec{x}$.
To show 3, 
consider two distinct equivalence classes $\ec{y}$ and $\ec{z}$,
and assume by contradiction that $\nf{(\Val_1 \sigma)} \ec{y} \aequiv \nf{(\Val_1 \sigma)} \ec{z}$.
Due to statement 3 for $\Val_1$,
$\ec{y}$ and $\ec{z}$ must have (minimal) subterms 
where $t_1$ occurs in $\Val_1 \ec{y}$ the same position (call it $p$) as $t_2$ occurs in $\Val_1 \ec{z}$,
and $t_1 \not\aequiv t_2$.
%We first show that $\nf{(\Val_1 \sigma)} \ec{y}$ (and for the same reasons, $\nf{(\Val_1 \sigma)} \ec{z}$)
%also contains position $p$.
%If this were not the case,
%then without loss of generality we may assume
%$\Val_1 \sigma \ec{y}$ contains a strict bisimilar subterm 
%$\mu {\Varec{w}}. u$ that strictly contains $t_1 \sigma$.
If $t_1$ (resp. $t_2$) is a free variable that is not ${\Varec{x}}$,
then $\nf{(\Val_1 \sigma)} \ec{y}$ (resp. $\nf{(\Val_1 \sigma)} \ec{z}$)
contains $t_1$ (resp. $t_2$) at position $p$,
and $\nf{(\Val_1 \sigma)} \ec{z}$ (resp. $\nf{(\Val_1 \sigma)} \ec{y}$) does not.
If $t_1$ is of the form $\mu w_1. \const{C}( \vec t )$ 
and $t_2$ is of the form $\mu w_2. \const{D}( \vec u )$,
then the expansion \rem{define} of $\nf{(\Val_1 \sigma)} \ec{y}$ and $\nf{(\Val_1 \sigma)} \ec{z}$
are $\alpha$-disequivalent at position $p$.
Since $t_1$ and $t_2$ are minimal,
say $t_1$ is of the form $\mu w. \const{C}( \vec t )$,
and $t_2$ is ${\Varec{x}}$.
Since $\sigma$ maps ${\Varec{x}}$ to a closed $\mu$-term $\mu {\Varec{x}}. t$,
we have that $FV( t_1 ) \subseteq \{ {\Varec{x}} \}$,
or else the expansion of $\nf{(\Val_1 \sigma)} \ec{y}$ and $\nf{(\Val_1 \sigma)} \ec{z}$
are $\alpha$-disequivalent at position $p$ since they do not contain the same free variables.
Since $\Val_1 \ec{y} \interp{t_1}$ is normal,
there is a closed $\mu$-term $v \in \mathcal{T}^\M_{\Varec{x}}( \Val_1 )$
such that $v \vsimv{\Varec{x}} \Val_1 \ec{y} \interp{t_1}$.
Thus, by assumption on our selection of $\mu {\Varec{x}}. t$,
we have $\mu {\Varec{x}}. t \not\vsimv{\Varec{x}} \Val_1 \ec{y} \interp{t_1}$,
which implies that the expansion of $\nf{(\Val_1 \sigma)} \ec{y}$ and $\nf{(\Val_1 \sigma)} \ec{z}$
are $\alpha$-disequivalent at position $p$. \rem{more detail}

>>>>

\hyphenation{data-type data-types co-data-type co-data-types isa-belle sledge-hammer
non-redun-dant}

@@

\paragraph{Correctness.}
An application of a rule is \emph{redundant} if at least one branch in the
conclusion does not add a new equality to $\Ec$ and is not $\bot$.
A \emph{derivation tree} is a tree whose nodes are finite sets of equalities and
where each non-root node is obtained by
a nonredundant application of a derivation rule to its parent node.
A derivation tree is \emph{closed} if all of its leaf nodes are $\bot$.
A node is \emph{saturated} if no nonredundant instance of a rule can be applied to it.

If there exists a closed derivation tree with root node $\Ec$, we conclude that $\Ec$ is $\thD$-unsatisfiable.
And if there exists a derivation tree with root node $\Ec$ that contains a saturated node, we conclude that $\Ec$ is $\thD$-satisfiable.

@@

We now define the class of interpretations for $\thD$.
In the following, $\M( \tau )$ denotes the interpretation type $\tau$ in $\M$,
that is, a set of domain elements for that type.
For a term $t$, $\M( t )$ denotes the interpretation of $t$ in $\M$,
that is, an element of $\M( \tau )$ where $\tau$ is the type of $t$.
In the following,
\emph{all} types are interpretated as a set of $\mu$-terms.
This is done to simplify the presentation.
Since all values of types in $\Data$ and $\Nondata$ are acyclic,
a typical representation for these values can be obtained by
dropping each of its $\mu$-bindings. 
For example, the $\mu$-term $\MU x.\; \const{0}$ representing the integer constant zero
is simply $\const{0}$ after dropping the $\mu$-binding of $x$.

@@

We now show that if statements 2 and 3 hold for some $\Val_1$,
then 2 and 3 also hold for $\nf{(\Val_1 \sigma)}$,
where $\sigma$ is a substitution of the form $\{ {\Varec{x}} \mapsto \MU{\Varec{x}}.\; t \}$,
${\Varec{x}} \in FV( \Val_1 )$,
and $\MU {\Varec{x}}.\; t$ is not $\alpha$-equivalent to a term in $\mathcal{T}^\M_{\Varec{x}}(\Val_1)$.
To show 2,
by assumption of 2 on $\Val_1$, we have $\Val_1 \ec{y} \interp{ u } = \Val_1 \ec{x}$ for
all subterms $u$ of $\Val_1 \ec{y}$ where $\muvar(t) = {\Varec{x}}$.
Thus, $\Val_1 \sigma \ec{y} \interp{ u } = \Val_1 \sigma \ec{x}$ and
$\nf{(\Val_1 \sigma)} \ec{y} \interp{ u } = \nf{(\Val_1 \sigma)} \ec{x}$.
To show 3, 
consider two distinct equivalence classes $\ec{y}$ and $\ec{z}$,
and assume by contradiction that $\nf{(\Val_1 \sigma)} \ec{y} \aequiv \nf{(\Val_1 \sigma)} \ec{z}$.
Due to statement 3 for $\Val_1$,
$\ec{y}$ and $\ec{z}$ must have (minimal) subterms 
where $t_1$ occurs in $\Val_1 \ec{y}$ the same position (call it $p$) as $t_2$ occurs in $\Val_1 \ec{z}$,
and $t_1 \not\aequiv t_2$.
%We first show that $\nf{(\Val_1 \sigma)} \ec{y}$ (and for the same reasons, $\nf{(\Val_1 \sigma)} \ec{z}$)
%also contains position $p$.
%If this were not the case,
%then without loss of generality we may assume
%$\Val_1 \sigma \ec{y}$ contains a strict bisimilar subterm 
%$\MU {\Varec{w}}.\; u$ that strictly contains $t_1 \sigma$.
If $t_1$ (resp. $t_2$) is a free variable that is not ${\Varec{x}}$,
then $\nf{(\Val_1 \sigma)} \ec{y}$ (resp. $\nf{(\Val_1 \sigma)} \ec{z}$)
contains $t_1$ (resp. $t_2$) at position $p$,
and $\nf{(\Val_1 \sigma)} \ec{z}$ (resp. $\nf{(\Val_1 \sigma)} \ec{y}$) does not.
If $t_1$ is of the form $\MU w_1.\; \const{C}( \vec t )$ 
and $t_2$ is of the form $\MU w_2.\; \const{D}( \vec u )$,
then the expansion \rem{define} of $\nf{(\Val_1 \sigma)} \ec{y}$ and $\nf{(\Val_1 \sigma)} \ec{z}$
are $\alpha$-disequivalent at position $p$.
Since $t_1$ and $t_2$ are minimal,
say $t_1$ is of the form $\MU w.\; \const{C}( \vec t )$,
and $t_2$ is ${\Varec{x}}$.
Since $\sigma$ maps ${\Varec{x}}$ to a closed $\mu$-term $\MU {\Varec{x}}.\; t$,
we have that $FV( t_1 ) \subseteq \{ {\Varec{x}} \}$,
or else the expansion of $\nf{(\Val_1 \sigma)} \ec{y}$ and $\nf{(\Val_1 \sigma)} \ec{z}$
are $\alpha$-disequivalent at position $p$ since they do not contain the same free variables.
Since $\Val_1 \ec{y} \interp{t_1}$ is normal,
there is a closed $\mu$-term $v \in \mathcal{T}^\M_{\Varec{x}}( \Val_1 )$
such that $v \vsimv{\Varec{x}} \Val_1 \ec{y} \interp{t_1}$.
Thus, by assumption on our selection of $\MU {\Varec{x}}.\; t$,
we have $\MU {\Varec{x}}.\; t \not\vsimv{\Varec{x}} \Val_1 \ec{y} \interp{t_1}$,
which implies that the expansion of $\nf{(\Val_1 \sigma)} \ec{y}$ and $\nf{(\Val_1 \sigma)} \ec{z}$
are $\alpha$-disequivalent at position $p$. \rem{more detail}
